\chapter{FUTURE ENCHANCEMENT}
\section{Unfulfilled Objectives and Remaining Challenges}

Despite the successful implementation of real-time emotion recognition from voice, several objectives remain partially addressed:
\begin{itemize}
    \item \textbf{Contextual and Temporal Awareness:} 
    The system analyzes short audio segments independently, without fully considering the flow of conversation or gradual emotional changes over time.
    
    \item \textbf{Multi-Speaker Handling:} 
    Interactions involving multiple speakers, overlapping voice, or rapid exchanges are not fully supported.
    
    \item \textbf{Emotion Intensity Levels:} 
    While emotions are classified into categories, the system does not quantify intensity, limiting its ability to distinguish subtle versus strong emotional expressions.
    
    \item \textbf{Language and Cultural Scope:} 
    The model primarily supports English but has limited Nepali support. Performance may still be restricted for other languages or culturally diverse emotional expressions.
    
    \item \textbf{Multimodal Limitations:} 
    Only voice-based features are analyzed; visual cues, textual content, or physiological signals are not incorporated, which could further enhance accuracy.
\end{itemize}

\section{Planned Improvements}

To further enhance the system, the following directions are following: 

\begin{itemize}
    \item \textbf{Dataset Expansion:} 
    Incorporate more diverse speakers, languages (including Nepali and others), and spontaneous conversational recordings to improve generalization and reduce bias.
    
    \item \textbf{Model Refinement:} 
    Introduce temporal modeling and attention mechanisms to capture context and highlight emotionally salient segments.
    
    \item \textbf{Feature Optimization:} 
    Explore additional acoustic features, such as formant frequencies and prosodic patterns, alongside advanced data augmentation to improve robustness.
    
    \item \textbf{Practical Enhancements:} 
    Extend system capabilities to handle multi-speaker scenarios and provide richer insights on emotion intensity for more nuanced feedback.
\end{itemize}

